---
output: word_document
---
# Preditive model for Weight Lifting Style using Accelerometer dataset

## Abstract
This is produced as required by Coursera Practical Machine Learning Assignment.  
Human Activity Recognition has emerged as a new key research due to the wide availability of wearable sensons and accelerometers from Jawbone Up, Nike FuelBand and Fitbit. The dataset collected from accelerometers on the belt, forearm, arm adn dumbell of 6 people. 

In this paper we try to develop a predictive model to predict the manner in which they did the exercise. This is the "classe" variable in the training set.

## Data Loading
The pml-training was downloaded from https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv and saved in the project directory
The pml-testing was downloaded from https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv and saved in the project directory

```{r}
## read the training dataset 
raw_dataset = read.csv('pml-training.csv')
## read the validation dataset 
pml_testing = read.csv('pml-testing.csv')

```
The pml-training dataset has `r nrow(raw_dataset)` rows and `r ncol(raw_dataset)` variables/columns

## Data Preparation for modeling

Partion the raw_dataset into two parts - one for training the model and the other for testing model

```{r}
library(caret)
library(randomForest)

#summary(raw_dataset)
set.seed(5678)
trainIndex = createDataPartition(raw_dataset$classe, list=FALSE, p=0.7)
train_set = raw_dataset[trainIndex,]
test_set  = raw_dataset[-trainIndex,]
# since most of the stistic columns have no data, let us get rid of them
#goodCols = !sapply(strsplit(names(train_set),"_"), function(x) x[1]) %in% c("kurtosis","skewness","max","min","amplitude","var","avg","stddev")
#train_set = train_set[,goodCols]
#test_set = test_set[,goodCols]
#summary(train_set)

```

Remove all indicators with near zero variance as they are not useful in modeling. After that remove any columns that are not numeric, but keep the classe varaible as this is the outcome variable 

```{r}
nzerovar = nearZeroVar(train_set)
train_set = train_set[-nzerovar]
test_set = test_set[-nzerovar]
pml_testing = pml_testing[-nzerovar]

numIndex = which(lapply(train_set,class) %in% c('numeric'))
train_set1 = train_set[,numIndex]
test_set1 = test_set[,numIndex]
pml_test1 = pml_testing[,numIndex]

```

Now, impute any missing values in the training data set

```{r}
preModel = preProcess(train_set1, method=c('knnImpute'))
ptrain_set = predict(preModel, train_set1)
ptest_set  = predict(preModel, test_set1)
pml_test_set = predict(preModel, pml_test1)

# add the classe column
ptrain_set = cbind(train_set$classe, ptrain_set)
ptest_set  = cbind(test_set$classe, ptest_set)
# need to fix the column name
names(ptrain_set)[1] = "classe"
names(ptest_set)[1] = "classe"

```
# Model Building

Using the "caret" package, build a random forest model, since computationally intensive will use optimized parameters
Then check the accuracy of the model for in-sample data
```{r}
RFmodel = randomForest(classe ~ ., ptrain_set, ntree=500, mtry=32)

# check accuracy of model for in-sample data
train_pred = predict(RFmodel, ptrain_set)
confusionMatrix(train_pred, ptrain_set$classe)

```

As we can see that the accuracy for the model is perfect for training data, which could sometimes indicate an overfit based on the training dataset. This can be verified by checking the accuracy for test data

```{r}
test_pred = predict(RFmodel, ptest_set)
confusionMatrix(test_pred, ptest_set$classe)

```
## Validation of the Model

As we can see from the results that the accuracy of the model is close to 99%, let us use the model to predict for the sample test data set provided as part of the assignment

```{r}
results = predict(RFmodel, pml_test_set)
print("The results:")
results
#confusionMatrix(results, pml_test_set$classe)

```

## Conclusion
The model built using the caret and randomForest packages for the given dataset seem to work very well as it predicted with good accuracy for both test data as well as the validation data. 


##References:
Weight Lifting Dataset is provided as part of Human Activity Recognition in the paper referred below
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.
